Spatial Reasoning and Data Displays

### 2015-03-30 Paper/InfoVis2015/TestingVisualAptitude.Rnw
switched out biplot by ggbiplot

@

@@

<<biplot-pca4, fig.cap="", fig.width=4.5, fig.height=4.5, warning=F, message=F, out.width='.45\\linewidth', comment="|", fig.cap="Biplots of principal components 1-4 with observations. Principal component analysis was performed on the four cognitive tests used to understand the association between the cognitive skills required for these tests and the skills required for the lineup protocol.  \\label{fig:biplots4}", echo=F, fig.env="figure">>=
par(mar=c(4,4,.1,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
biplot(pca2, choices=1:2, pc.biplot=T, cex=c(.5, 1), xlim=c(-2.5,2.5), ylim=c(-2.5,2.5))
biplot(pca2, choices=3:4, pc.biplot=T, cex=c(.5, 1), xlim=c(-3,3), ylim=c(-2.5,2.5))
@
Figure~\ref{fig:biplots4} shows that the first PC does not differentiate between any of the tasks; it might be best understood as a general aptitude factor. All of the remaining principal components distinguish between the cognitive tasks; we can utilize this separation by incorporating the lineup task and examining changes in the principal components.

@@

<<biplots-pca5, fig.cap="", fig.width=4.5, fig.height=4.5, warning=F, message=F, out.width='.45\\linewidth', comment="|", fig.cap="Biplots of principal components 2-5 with observations. The lineup task appears to be most similar to the figure classification task, based on the plot of PC2 vs. PC3.  \\label{fig:biplots5}", echo=F, fig.env="figure">>=
par(mar=c(4,4,.1,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
biplot(pca, choices=2:3, pc.biplot=T, cex=c(.5, 1), xlim=c(-2.5,2.5), ylim=c(-2.5,2.5))
biplot(pca, choices=4:5, pc.biplot=T, cex=c(.5, 1), xlim=c(-2,2), ylim=c(-2,2))
@

Biplots \cite{biplots:gabriel} of the remaining components are provided in figure~\ref{fig:biplots5}. Figure classification is strongly related to lineups (PC2, PC3). Performance on the visual search task  is also related to lineup performance (PC3). These two components highlight the shared demands of the lineup task and the figure classification task: participants must establish categories from provided stimuli and then classify the stimuli accordingly.

>>>>

ggbiplot <- function(pca, choices, segColor="black") {
  u <- pca$x[,choices]
  v <- pca$rotation[,choices]
  u <- data.frame(sweep(u, MARGIN=2, 1/pca$sdev[choices], FUN="*"))
  v <- data.frame(sweep(v, MARGIN=2, 1/sqrt(rowSums(v^2))[choices], FUN="*"))
  v$label <- row.names(v)

  segColor <- rep(segColor, length=nrow(v))

  qplot(u[,1], u[,2], data=u, colour=I("grey50"), shape=I(1), size=I(3)) + theme_bw() + 
    theme(aspect.ratio=1) + coord_equal() + 
  geom_segment(aes(x=0, y=0, xend=v[,1], yend=v[,2], colour=segColor), data=v, arrow=arrow(length = unit(0.3, "cm"))) + 
    xlab(sprintf("PC%d",choices[1])) + ylab(sprintf("PC%d",choices[2])) +
    geom_text(aes(x=v[,1], y=v[,2], label=label, colour=segColor, 
                  hjust=0.15*(0.5-sign(v[,1])), vjust=0.5-sign(v[,2])), data=v) +
    scale_colour_identity()
}
@

@@

<<biplot-pca4, fig.cap="", fig.width=4.5, fig.height=4.5, warning=F, message=F, out.width='.45\\linewidth', comment="|", fig.cap="Biplots of principal components 1-4 with observations. Principal component analysis was performed on the four cognitive tests used to understand the association between the cognitive skills required for these tests and the skills required for the lineup protocol.  \\label{fig:biplots4}", echo=F, fig.env="figure">>=
#par(mar=c(4,4,.1,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
dimnames(pca2$rotation)[[1]] <- c("card rotation", "figure classification", "paper folding", "visual search")

ggbiplot(pca2, choices=1:2) + xlim(c(-2.5,2.5)) +  ylim(c(-2.5,2.5))
ggbiplot(pca2, choices=3:4) + xlim(c(-3,3)) + ylim(c(-2.5,2.5))
@
Figure~\ref{fig:biplots4} shows that the first PC does not differentiate between any of the tasks; it might be best understood as a general aptitude factor. All of the remaining principal components distinguish between the cognitive tasks; we can utilize this separation by incorporating the lineup task and examining changes in the principal components.

@@

% <<biplots-pca5, fig.cap="", fig.width=4.5, fig.height=4.5, warning=F, message=F, out.width='.45\\linewidth', comment="|", fig.cap="Biplots of principal components 2-5 with observations. The lineup task appears to be most similar to the figure classification task, based on the plot of PC2 vs. PC3.  \\label{fig:biplots5}", echo=F, fig.env="figure">>=
% par(mar=c(4,4,.1,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
% biplot(pca, choices=2:3, pc.biplot=T, cex=c(.5, 1), xlim=c(-2.5,2.5), ylim=c(-2.5,2.5))
% biplot(pca, choices=4:5, pc.biplot=T, cex=c(.5, 1), xlim=c(-2,2), ylim=c(-2,2))
% @

<<biplots-pca5, fig.cap="", fig.width=4.5, fig.height=4.5, warning=F, message=F, out.width='.45\\linewidth', comment="|", fig.cap="Biplots of principal components 2-5 with observations. The lineup task appears to be most similar to the figure classification task, based on the plot of PC2 vs. PC3.  \\label{fig:biplots5}", echo=F, fig.env="figure">>=

dimnames(pca$rotation)[[1]] <- c("lineup", "card rotation", "figure classification", "folding", "visual search")
ggbiplot(pca, choices=2:3, segColor=c("red", rep("black", 4))) + xlim(c(-2.6,2.6)) +  ylim(c(-2.6,2.6))
ggbiplot(pca, choices=4:5, segColor=c("#FF3333", rep("black", 4)))
@
Biplots \cite{biplots:gabriel} of the remaining components are provided in figure~\ref{fig:biplots5}. Figure classification is strongly related to lineups (PC2, PC3). Performance on the visual search task  is also related to lineup performance (PC3). These two components highlight the shared demands of the lineup task and the figure classification task: participants must establish categories from provided stimuli and then classify the stimuli accordingly.

### 2015-03-30 Paper/InfoVis2015/TestingVisualAptitude.tex
switched out biplot by ggbiplot

\documentclass[journal]{vgtc}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}                % final (journal style)
%\documentclass[review,journal]{vgtc}         % review (journal style)
%\documentclass[widereview]{vgtc}             % wide-spaced review
%\documentclass[preprint,journal]{vgtc}       % preprint (journal style)
%\documentclass[electronic,journal]{vgtc}     % electronic version, journal

@@

%% Uncomment below to include a teaser figure.
   \teaser{
    \centering
\begin{minipage}[c]{.5\textwidth}
\includegraphics[width=\textwidth]{lineup}
\end{minipage}
\begin{minipage}[c]{.3\textwidth}
    \includegraphics[height=.8\textwidth]{Figure/fig-biplots-pca5-1}
    \includegraphics[height=.8\textwidth]{Figure/fig-biplots-pca5-2}
\end{minipage}
    \caption[Sample Lineup]{``Which of these plots is the most different from the others?"\newline
    This is the question participants are asked to answer for each of the lineups in the study. In the figure, a sample lineup of boxplots is shown. %Participants are instructed to choose the plot which appears most different from the others. 
    Panel 4 is the target plot, because the two groups have a large difference in medians.\label{fig:lineup}}
  }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%% START OF THE PAPER %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Sat Mar 28 22:35:16 2015
\begin{table}[htb]
\centering
\caption{Importance of principal components in an analysis of four tests of spatial ability: figure classification, paper folding, card rotation, and visual search.\label{tab:PCAvariance4}} 
{\footnotesize
\begin{tabular}{rrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 \\ 
  \hline
Standard deviation & 1.61 & 0.81 & 0.73 & 0.49 \\ 
  Proportion of Variance & 0.64 & 0.16 & 0.13 & 0.06 \\ 
  Cumulative Proportion & 0.64 & 0.81 & 0.94 & 1.00 \\ 
   \hline
\end{tabular}
}
\end{table}

@@

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Sat Mar 28 22:35:16 2015
\begin{table}[htb]
\centering
\caption{Rotation matrix for principal component analysis of the four cognitive tests (visual search, paper folding, card rotation, figure classification).\label{tab:PCArotation4}} 
\begin{tabular}{rrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 \\ 
  \hline
card.rot & 0.55 & -0.19 & -0.38 & 0.72 \\ 
  fig.class & 0.46 & 0.58 & 0.66 & 0.14 \\ 
  folding & 0.52 & 0.33 & -0.53 & -0.59 \\ 
  vis.search & 0.46 & -0.72 & 0.38 & -0.34 \\ 
   \hline
\end{tabular}
\end{table}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}[]

{\centering \includegraphics[width=.45\linewidth]{Figure/fig-biplot-pca4-1} 
\includegraphics[width=.45\linewidth]{Figure/fig-biplot-pca4-2}

}

\caption{Biplots of principal components 1-4 with observations. Principal component analysis was performed on the four cognitive tests used to understand the association between the cognitive skills required for these tests and the skills required for the lineup protocol.  \label{fig:biplots4}\label{fig:biplot-pca4}}
\end{figure}

\end{knitrout}
Figure~\ref{fig:biplots4} shows that the first PC does not differentiate between any of the tasks; it might be best understood as a general aptitude factor. All of the remaining principal components distinguish between the cognitive tasks; we can utilize this separation by incorporating the lineup task and examining changes in the principal components.

Next, we add the lineup results to our principal components analysis. The results of the principal component analysis of the four cognitive tests and lineup results are similar to the results of the cognitive tests alone; this suggests that lineups require at least some of the skills measured by the other cognitive tests. Table~\ref{tab:PCAvariance5} shows the importance of each principal component; from the distribution of the variance components, we see that the lineup test spans an additional dimension within the space of the four established tests. 
% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Sat Mar 28 22:35:16 2015
\begin{table}[ht]
\centering
\caption{Importance of principal components, analyzing all five tests.\label{tab:PCAvariance5}} 
{\footnotesize
\begin{tabular}{rrrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 \\ 
  \hline
Standard deviation & 1.73 & 0.84 & 0.75 & 0.70 & 0.48 \\ 
  Proportion of Variance & 0.60 & 0.14 & 0.11 & 0.10 & 0.05 \\ 
  Cumulative Proportion & 0.60 & 0.74 & 0.85 & 0.95 & 1.00 \\ 
   \hline
\end{tabular}
}
\end{table}

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Sat Mar 28 22:35:16 2015
\begin{table}[ht]
\centering
\caption{PCA Rotation matrix for all five tests. The first principal component is essentially an average of all five tests. 
             \label{tab:PCArotation5}} 
\begin{tabular}{rrrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 \\ 
  \hline
lineup & 0.42 & 0.49 & -0.46 & 0.60 & -0.10 \\ 
  card.rot & 0.50 & -0.30 & 0.28 & 0.23 & 0.73 \\ 
  fig.class & 0.43 & 0.45 & -0.15 & -0.75 & 0.18 \\ 
  folding & 0.47 & 0.07 & 0.68 & 0.04 & -0.56 \\ 
  vis.search & 0.41 & -0.69 & -0.48 & -0.15 & -0.33 \\ 
   \hline
\end{tabular}
\end{table}

From the rotation matrix (see Table~\ref{tab:PCArotation5}) we see that  
the first principal component, PC1, is again essentially an average across all tests and accounts for 60.1\% of the variance in the data.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}[]

{\centering \includegraphics[width=.45\linewidth]{Figure/fig-biplots-pca5-1} 
\includegraphics[width=.45\linewidth]{Figure/fig-biplots-pca5-2}

}

\caption{Biplots of principal components 2-5 with observations. The lineup task appears to be most similar to the figure classification task, based on the plot of PC2 vs. PC3.  \label{fig:biplots5}\label{fig:biplots-pca5}}
\end{figure}

\end{knitrout}

Biplots \cite{biplots:gabriel} of the remaining components are provided in figure~\ref{fig:biplots5}. Figure classification is strongly related to lineups (PC2, PC3). Performance on the visual search task  is also related to lineup performance (PC3). These two components highlight the shared demands of the lineup task and the figure classification task: participants must establish categories from provided stimuli and then classify the stimuli accordingly.

The visual search task is also clearly important to lineup performance: PC3 captures the similarity between the visual search and lineup performance, and aspects of these tasks are negatively correlated with aspects of the paper folding and card rotation tasks within PC3. Paper folding does not seem to be strongly associated with lineup performance outside of the first principal component; card rotation is only positively associated with lineup performance in PC4.

PC4 captures the similarity between lineups and the card rotation task and separates this similarity from the figure classification task; this similarity does not account for much extra variance (10\%), but it may be that only some lineups require spatial rotation skills. PC5 contains only 5\% of the remaining variance, and is thus not of much interest, however, it seems to capture the relationship between the card rotation task and the paper folding and visual search tasks.

Figure classification is strongly related to lineups, and as in the four-component PCA, figure classification is strongly represented in the first two principal components. While lineups do span a separate dimension, the PCA suggests that they are most closely related to the figure classification task, and least related to the visual searching task.

This emphasizes the underpinnings of lineups: the test utilizes a visual medium, but is ultimately are a classification task presented in a graphical manner. Using lineups as a proxy for statistical significance tests is similar to using a classifier on pictoral data: while the data is presented ``graphically", the participant is actually classifying the data based on underlying summary statistics.

\subsection{Linear model of demographic factors}
Note that all of the demographic variables in the survey are highy correlated, for example there is a high correlation between STEM majors and taking calculus.
Similarly, the correlation  between having taken a statistics class and having been involved in mathematics or statistics research is high. Only one student is doing research who has not taken a statistics course.

A principal component of the five math/stats questions splits the variables into two main areas: the first principal component is an average of math skills, calculus 1 and STEM, while the second principal component is an average of having taken a statistics class and doing research. We therefore decided to use sums of these variables to come up with a separate math and a stats score. Note, that the correlation between the math and the stats score is almost zero.

We fit a linear model of lineup scores in the thus modified demographic variables and the test scores from the visuo-spatial tests, selecting the best model using AIC and stepwise backwards selection. The result is shown in Table~\ref{tab:m1}. Only two covariates stay in the model: PC1 and MATH, reflecting two dimensions of what affects lineup scores. We can think of PC1 as a measure of innate visual or intellectual ability, while the MATH score is a matter of both ability and training. The remaining principal components were not sufficiently associated with lineup score to be included in the model.

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Sat Mar 28 22:35:16 2015
\begin{table}[ht]
\centering
\caption{Estimates and significances of a linear model of lineup scores.} 
\label{tab:m1}
\begin{tabular}{rrrrr}
  \hline
 & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
  \hline
(Intercept) & 14.1192 & 1.9149 & 7.37 & 0.0000 \\ 
  PC1 & 1.7672 & 0.5230 & 3.38 & 0.0018 \\ 
  MATH & 2.1246 & 0.8732 & 2.43 & 0.0202 \\ 
   \hline
\end{tabular}
\end{table}

%\afterpage{\clearpage}
%
\subsection{Lineup Types}
Each of the three sets of 20 lineups was taken from previous experiments examining which type of plot most effectively conveyed important characteristics of the underlying data set. 
\begin{figure*}[htbp]\centering
\subfloat[Set 1: Boxplot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/boxplot_1_15_5_4_13}}
\subfloat[Set 1: Density plot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/density_1_15_5_4_13}}
\subfloat[Set 1: Dotplot ]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/dotplot_1_15_5_4_13}}
\subfloat[Set 1: Histogram]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/histogram_1_15_5_4_13}}\\
\subfloat[Set 2: Boxplot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f94f6965d7}}
\subfloat[Set 2: Boxplot with jittered points]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f9e65f0e4}}
\subfloat[Set 2: Bee swarm boxplot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f9fc58d73}}
\subfloat[Set 2: Violin plot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f93acbb7e4}}\\
\subfloat[Set 3: QQ-plot with guide line]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups3/file13876a225d34-single}}
\subfloat[Set 3: QQ-plot with acceptance bands]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups3/file13877c93dc78-single}}
\subfloat[Set 3: QQ-plot rotated 45 degrees.]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups3/file13878e5c92f-single}}
\caption{Lineup plot types used in the experiment. \label{fig:lineups}}
\end{figure*}

\paragraph{Lineup Set 1}
The experiment in the first lineup section examined the use of boxplots, density plots, histograms, and dotplots to compare two groups which vary in mean and sample size. This set of lineups consists of 20 plots selected from the plots used in the full experiment; each set of data is displayed with each of the four plot types. Figure \ref{fig:lineups} (a)-(d) provides an example of the four plot types displaying the same data.

\paragraph{Lineup Set 2}
The second lineup section also explored two groups of data, this time comparing boxplots, bee swarm boxplots, boxplots with overlaid jittered data, and violin plots, as shown in figure \ref{fig:lineups} (e)-(h). Participants were much more accurate in this experiment than in the experiment described previously, because of the types of plots compared as well as the underlying data distributions.

\paragraph{Lineup Set 3}
The final lineup set explored QQ-plots from various model simulations, using reference lines, acceptance bands, and rotation to determine which plots allowed participants to most effectively identify violations of normality. Rotated QQ-plots showed lower performance because participants were able to more accurately compare acceptance bands to residuals, and thus could identify that the reference bands were too liberal. As a result, performance was somewhat lower for rotated plots, even though participants were more accurate when comparing the residuals to the reference bands.

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{Figure/fig-LineupTypeSPM-1}
\caption{Pairwise scatterplots of test scores, with lineups separated into the three lineup tasks. All lineup tasks are moderately correlated with the figure classification task, and while tasks 1 and 2 are most strongly correlated with figure classification, lineup task 3 is most strongly correlated with the card rotation and paper folding tasks. \label{fig:lineuptaskmatrix}}
\end{figure}
Figure~\ref{fig:lineuptaskmatrix} shows the correlations between the three lineup tasks and the figure classification, card rotation, and paper folding tasks. The visual search task is only slightly correlated with the three lineup tasks and is therefore omitted from this figure. Performance on lineup tasks 1 and 2, which dealt with the distributuion of two groups of numerical data, is most strongly correlated with the performance on the figure classification task, which measures general reasoning ability.
Performance on lineup task 3, which investigated the potential to visually identify nonnormality in residual QQ-plots, is more associated with the card rotation and paper folding tests, which measure visuospatial ability. This suggests that certain lineup tasks may require more visual ability than others; in the case of the QQ-lineups a successful evaluation  needed  participants to mentally rotate plots to compare vertical distances, requiring more mental manipulation than the first two lineup tasks.

In order to examine which lineup tasks are most closely associated with visual abilities tested in the aptitude portions of an experiment, we employ principal component analysis on participant scores averaged across each block of lineups.

Principal component analysis separates multivariate data into orthogonal components using a rotation matrix to transform correlated input data into an orthogonal space. The importance of a principal component is also evaluated to assess the proportion of the overall variance in the data contained within the component (Table \ref{tab:PCAvariance5} shows the importance of each PC for the analysis of the aggregate lineup score and the four aptitude tests).

\begin{figure}[ht]
\includegraphics[width=\linewidth]{Figure/fig-lineupblockpca-1}
\caption{Principal Component influence plot, showing each input variable's contribution to the principal component, scaled by the proportion of variability in the data contained in the principal component.\label{fig:influence}}
\end{figure}

For variables $X_i$, $i=1, ..., K$ a principal component analysis results in a set of $K$ principal  components $PC_j$, $j=1, ..., K$, given as the rows of the rotation matrix, $M$ (which has dimension $K\times K$, indexed by $i$ and $j$ respectively).

The importance $I_j$ of each component is determined by the amount of variability the data exhibits along each of the principal axis.

The influence of variable $X_i$ on component $PC_j$ is then defined as:
%
$$\text{Influence of variable} X_i \text{ on PC}_j = M_{ij}\times I_j$$
%
Influence is therefore a measure of the contribution of an input variable to the principal component, scaled by the importance of that principal component to the overall variability in the data. Small (absolute) values indicate that there is little influence of $X_i$ on $PC_j$; negative values show the direction of the influence (to maintain the separation of variables as part of principal components analysis).

Figure~\ref{fig:influence} shows the influence of each input variable on each principal component for a principal component analysis  including each lineup task as a separate input variable. The input variables are shown on the $y$ axis, with the influence of the variable on each PC shown on the $x$ axis. This allows us to consider the rotation matrix visually (while accounting for the importance of each principal component); for instance, we see that again, the first PC accounts for most of the variance and seems to represent general visual aptitude.

PC2 emphasizes the overlapping variation in performance on lineup test 1 and the figure classification test. PC3 emphasizes the additional variation in performance on lineup test 3 and the visual search test, while PC4 emphasizes the extra variability in performance on lineup test 2 and the paper folding test. All three lineups, plus the figure classification, paper folding, and visual search tests contribute to PC5. PC6 and PC7 jointly account for less than 10\% of the variance in the data and do not display any distinct patterns in the loadings.

While lineups constitute a distinct principal component when aggregated into a single score, this PCA of the separate lineup types and the cognitive tests indicates that different lineup experiments exist in different principal component loadings. Overall, there is an additional principal component gained from separating the lineup blocks by experiment type. As lineup tasks 1 and 2 contained similar plot types, it is possible that those two tasks overlap in the component space while lineup task 3 is distinct.

The relationship between participant performance on different types of lineups (and different types of plots) and performance on tests of spatial ability bears further investigation; this study suggests that there may be an effect, but there is simply not enough variation to make definitive conclusions about the relationship between different measures of visual aptitude and performance on specific lineup tasks.

A larger study might not only address the question of which lineup tasks require certain visual skills, but also the use of different types of plots from a perceptual perspective. Preliminary results of performance on different types of plots are shown in Appendix~\ref{app:plottypes}, but a larger study is needed for definitive results. The advantage of the lineup protocol is that it allows us to not only consider individual performance but also to compare aggregate performance on different types of plots. Integrating information about the visual skills required for each type of plot provides information about the underlying perceptual skills and experience required to read different types of plots.

\section{Discussion and Conclusions}\label{sec:discussion}

Performance on lineups is strongly related to performance on tests of visual ability; however, this relationship is mediated by demographic factors such as major (STEM or not) and completion of calculus I.
In addition to these demographic factors, many facets of intelligence are highly correlated; participants who score higher on general aptitude tests may score higher on tests of visual ability (and may also score higher on lineup tests).

Despite these caveats, we have demonstrated that the general lineup task is most closely related to a classification task, rather than tests of spatial ability.
This is an important verification of a tool that is useful for examining statistical graphics, as it emphasizes the idea that while the testing medium is graphical in nature, the task is in fact a classification task, where the viewer must determine the most important features of each plot and then identify which plot is different.

When lineup tasks with different goals are viewed separately, there is some indication that different tasks are associated with different visual abilities.
Lineup tasks 1 and 2 are quite similar, and are more associated with the figure classification task; lineup task 3, while still moderately correlated with the figure classification task, is also moderately correlated with the visuospatial ability tests (paper folding, card rotation).
Future studies testing larger sets of lineups may be useful to understand which types of plots require additional visuospatial skills, as plots which appeal to a wider audience may be more successful when conveying important information.

In addition to this theoretical information, the figure classification test may be useful for pre-screening participants in future online lineup studies.
Such studies often suffer from participants who do not take the task seriously, and internal verification questions, as well as pre-qualification tasks are often used to reduce extraneous variability.
While it is impractical to require participants to score well on several different tests, it is reasonable to ask participants to pre-qualify for a task by completing a figure classification test.
As the figure classification test is different from the lineup task, this will not bias participants' scores on the domain of interest while ensuring that the participant pool is sufficiently motivated to complete the lineup questions.

The demographic results from this study indicate that in future lineup studies, it may be important to record information about participants' mathematical training, so that studies can be compared across participant pools with more reliability.

All results and data shown here were collected and analyzed in accordance with IRB \# 13-581.

\bibliographystyle{abbrv}
%%use following if all content of bibtex file should be shown
%\nocite{*}
\bibliography{references}

\end{document}

>>>>

\documentclass[journal]{vgtc}                % final (journal style)
%\documentclass[review,journal]{vgtc}         % review (journal style)
%\documentclass[widereview]{vgtc}             % wide-spaced review
%\documentclass[preprint,journal]{vgtc}       % preprint (journal style)
%\documentclass[electronic,journal]{vgtc}     % electronic version, journal

@@

%% Uncomment below to include a teaser figure.
   \teaser{
    \centering
\begin{minipage}[c]{.5\textwidth}
\includegraphics[width=\textwidth]{lineup}
\end{minipage}
\begin{minipage}[c]{.3\textwidth}
    \includegraphics[height=.8\textwidth]{Figure/fig-biplots-pca5-1}
    \includegraphics[height=.8\textwidth]{Figure/fig-biplots-pca5-2}
\end{minipage}
    \caption[Sample Lineup]{``Which of these plots is the most different from the others?"\newline
    This is the question participants are asked to answer for each of the lineups in the study. In the figure, a sample lineup of boxplots is shown. %Participants are instructed to choose the plot which appears most different from the others. 
    Panel 4 is the target plot, because the two groups have a large difference in medians.\label{fig:lineup}
    The plots on the right show biplots of principal components 2-5 with observations. The lineup task appears to be most similar to the figure classification task, based on the plot of PC2 vs. PC3.}
  }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%% START OF THE PAPER %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

@@

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Mon Mar 30 14:32:39 2015
\begin{table}[htb]
\centering
\caption{Importance of principal components in an analysis of four tests of spatial ability: figure classification, paper folding, card rotation, and visual search.\label{tab:PCAvariance4}} 
{\footnotesize
\begin{tabular}{rrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 \\ 
  \hline
Standard deviation & 1.61 & 0.81 & 0.73 & 0.49 \\ 
  Proportion of Variance & 0.64 & 0.16 & 0.13 & 0.06 \\ 
  Cumulative Proportion & 0.64 & 0.81 & 0.94 & 1.00 \\ 
   \hline
\end{tabular}
}
\end{table}

@@

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Mon Mar 30 14:32:39 2015
\begin{table}[htb]
\centering
\caption{Rotation matrix for principal component analysis of the four cognitive tests (visual search, paper folding, card rotation, figure classification).\label{tab:PCArotation4}} 
\begin{tabular}{rrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 \\ 
  \hline
card.rot & 0.55 & -0.19 & -0.38 & 0.72 \\ 
  fig.class & 0.46 & 0.58 & 0.66 & 0.14 \\ 
  folding & 0.52 & 0.33 & -0.53 & -0.59 \\ 
  vis.search & 0.46 & -0.72 & 0.38 & -0.34 \\ 
   \hline
\end{tabular}
\end{table}
